---
title: "Ravens Stats Week 11"
author: "Yonathan Melamed"
date: '2023-11-16'
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nflfastR)
library(tidyverse)
library(knitr)
library(kableExtra)
pbp=load_pbp(season=2023)

pbp$pass_bucket=cut(pbp$air_yards, breaks = c(-10, 0, 10, 25, 100))
```

## Weakness in the Defense?

This is the first in a series of weekly articles I'll be writing that dive into some Ravens centered statistical analysis. Much of the data I will be 
using comes from nflFastR, an R package with play-by-play data.

```{r pressure,out.width= "65%", out.extra='style="float:right; padding:10px"', echo=FALSE, warning=FALSE, message=FALSE}
pbp %>% 
  filter(!is.na(aborted_play), !is.na(epa)) %>% 
  group_by(defteam, pass) %>% 
  summarise(avg_epa=mean(epa)) %>% 
  pivot_wider(names_from = pass, values_from = avg_epa) %>% 
  rename(rush='0', pass='1', team='defteam') %>% 
  ggplot2::ggplot(aes(x = pass, y = rush)) +
  ggplot2::geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3), alpha = .2) +
  #nflplotR::geom_mean_lines(aes(y0 = pass, x0 = rush)) +
  nflplotR::geom_nfl_logos(aes(team_abbr = team), width = 0.07, alpha = 0.7) +
  ggplot2::labs(
    x = "Passing EPA/play",
    y = "Rushing EPA/play",
    caption = "Data: @nflfastR",
    title = "2023 NFL Defensive EPA per Play"
  ) +
  scale_x_reverse() +
  ggplot2::theme_bw() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 12, hjust = 0.5, face = "bold")
  ) +
  ggplot2::scale_y_reverse() 

```

This week, I wanted to take a closer look at the Ravens defense. Mike McDonald has done a phenomenal job this year designing a cohesive unit across all levels. The front seven, which was thought to be the point of weakness going into the year, has exceeded expectations by a mile. McDonald has proven himself a genius at engineering pressures even without top tier talent (see Clowney, Van Noy, etc.). The linebackers have successfully replicated last year's performance post Roquan trade and are unquestionably among the best units in the league. Those avid Ravens fans who've watched the team flash defensive prowess only to succumb to elite quarterbacks came into the year worried about the lack of depth in the secondary (we've also unfortunately grown accustomed to watching the preseason CB-5 turn into the top option to guard players like Jamar Chase). Yet, the secondary has held strong, largely thanks to the development of Brandon Stephens and the solid play of free agents Ronald Darby, Rock Ya-Sin, and Arthur Maulet. Geno Stone has blossomed into a talented playmaker in the absence of Marcus Williams and Kyle Hamilton is developing into a real star. All this has translated into a defense that has given up the second least EPA in the NFL, which begs the question: Is there a weakness to the defense that may have not been exposed yet?



Diving into the data this week, I found a concerning trend that may show up this week. Watching the Browns game last week, I noticed multiple plays where the secondary seemed to have a play perfectly covered, only to give up a short underneath throw to the RB/TE. I decided to explore whether the defense was particular poor/stout at defending short/intermediate/deep passes. For the purposes of this analysis, I defined short passes as passes with 0-10 air yards, intermediate passes as 11-25 air yards, and deep passes over 25 air yards. I then calculated the EPA splits for each team in the league. What I found was mildly concerning:


```{r echo=FALSE, warning=FALSE, message=FALSE}
#defense
tot_plays = 
  pbp %>%  filter(pass_attempt==1, aborted_play==0, !is.na(air_yards)) %>% 
  group_by(defteam) %>% 
  summarise(tot_plays=n(), tot_epa=sum(epa)) 

pass_lengths = pbp %>% 
  filter(pass_attempt==1, aborted_play==0, !is.na(air_yards)) %>% 
  group_by(defteam, pass_bucket) %>% 
  summarise(EPA=sum(epa), plays=n()) %>% 
  left_join(tot_plays) %>% 
  mutate(per_plays=plays/tot_plays, per_epa=EPA/tot_epa) %>% 
  rename(Air_Yards=pass_bucket)

ravens=pass_lengths %>% filter(defteam=="BAL", !is.na(Air_Yards)) %>% select(defteam, Air_Yards, EPA, plays) 

kable(ravens, caption="**Ravens Defensive EPA by Air Yards (on thrown passes)**") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```


As you can see, the Ravens defense is excelling at preventing deep and intermediate passes (remember negative EPA is better when looking at defenses). This is pretty easy to see from watching this team. They don't give up big plays, and [that's an extremely important quality to have, as explored by PFF](https://www.pff.com/news/nfl-explosive-plays-and-re-thinking-offensive-success#:~:text=Explosive%20plays%20are%20overwhelming%20accrued,the%20opponent's%20talent%20and%20scheme.). However, we would be remiss to not consider this concern. In fact, as shown below, when we break down each team's EPA split by air yard bucket, the Ravens' short passing defense is among the worst in the league:


```{r echo=FALSE, warning=FALSE, message=FALSE}
#defense
tot_plays = 
  pbp %>%  filter(pass_attempt==1, aborted_play==0, !is.na(air_yards)) %>% 
  group_by(defteam) %>% 
  summarise(tot_plays=n(), tot_epa=sum(epa)) 

pass_lengths = pbp %>% 
  filter(pass_attempt==1, aborted_play==0, !is.na(air_yards)) %>% 
  group_by(defteam, pass_bucket) %>% 
  summarise(EPA=sum(epa), plays=n()) %>% 
  left_join(tot_plays) %>% 
  mutate(per_plays=plays/tot_plays, per_epa=EPA/tot_epa) %>% 
  rename(Air_Yards=pass_bucket)

league_defense=pass_lengths %>% filter(!is.na(Air_Yards)) %>% select(defteam, Air_Yards, EPA, plays) %>% arrange(-EPA)

kable(league_defense[1:10, ], caption="**League Defensive EPA by Air Yards (on thrown passes)**") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
    row_spec(9, bold = T, color = "white", background = "#D7261E")

```
Now, this is admittedly hard to explain. We can understand why teams are targeting the short passing game (quick pressure, solid coverage, etc.), but I'm curious as to why teams have seemingly been so succesfull at it. It does seem that when the Ravens do give up long drives, especially late in their 3 losses this year, they have been long, sustained drives built on the short passing game. But why? Roquan Smith has the top PFF coverage grade among linebackers, and although Patrick Queen's grades have slid in recent weeks, he still boosts a decent 60.7 grade. Meanwhile, Geno Stone and Kyle Hamilton both rank top 10 among safeties in coverage grades. 

So what gives? At this point, I haven't found an answer. Maybe it's schemed related, or maybe it's just noise. Should we even care? Interestingly enough, Thursday night's matchup may give us an answer. When we run the same analysis on league offenses, we see that Cincinnati has excelled at the short passing game. In fact, they have gained more EPA from short passes than any other team in the league.


```{r echo=FALSE, warning=FALSE, message=FALSE}
#offense
tot_plays = 
  pbp %>%  filter(pass_attempt==1, aborted_play==0, !is.na(air_yards)) %>% 
  group_by(posteam) %>% 
  summarise(tot_plays=n(), tot_epa=sum(epa)) 

pass_lengths = pbp %>% 
  filter(pass_attempt==1, aborted_play==0, !is.na(air_yards)) %>% 
  group_by(posteam, pass_bucket) %>% 
  summarise(EPA=sum(epa), plays=n()) %>% 
  left_join(tot_plays) %>% 
  mutate(per_plays=plays/tot_plays, per_epa=EPA/tot_epa) %>% 
  rename(Air_Yards=pass_bucket)

league_defense=pass_lengths %>% filter(!is.na(Air_Yards)) %>% select(posteam, Air_Yards, EPA, plays) %>% arrange(-EPA)

kable(league_defense[1:10, ], caption="**League Offensive EPA by Air Yards (on thrown passes)**") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),  position = "float_right") %>% 
    row_spec(5, bold = T, color = "white", background = "#1ED762")

```

In sum, the Ravens defense has seemingly been firing on all cylinders. However, we saw last week that they may have an Achilles heal. Thursday night's game against the Bengals will prove to be a litmus test for this defense as it squares off against an offensive prowess that has specifically excelled where the defense has quietly struggled. 

Thanks a lot for reading! In the future, I plan to include more in depth-analysis and creative visualizations, but wanted to get something out before today's game. Go Vens!
